The Natural History of Child Signals of Need in Utila, Honduras: An Exploratory Study

### 2024-10-22 Paper.qmd
Paper: worked on discussion.
Discussion: fixed typo.

As seen in Western cultures [@zeifman2013; @potegal2003; @koch2003] (check for others), the frequencies of crying and temper tantrums decreased with age. One possibility is that older children have greater ability to satisfy one's desires and behave in ways which match cultural expectations of teenagers and adults. Another is that since one's reproductive value increases as one survives periods of high mortality risk in childhood, the intensity of conflicts of interest with parents decreases since these children are likely to reach adulthood and require less investment to reach adulthood than younger children, current or future [@daly1995]. For this reason, less costly signals may be more likely to work to increase investment, all else being equal\[\[4\]\]. That said, all else may often not be equal if investment in older children is costlier, something which may be reflected in caretakers seeing older children as relatively needy. The decrease in crying and tantrum frequencies with age may also result from the potential for greater reputational costs, something which may also increase the credibility of such signals among older children as previously mentioned.

There was also a negative association between child sadness and age. However, interpretation of this finding is more difficult due to the fact that, unlike crying and temper tantrums, sadness is thought to have important non-signaling functions including drawing one’s attention to a problem, motivating deeper analysis of one’s situation, and causing withdrawal from situations with low fitness prospects [@nesse2019; @nettle2009; @allen2003; @hagen2011]. By asking about sadness as a whole, our measure was unable to disentangle information caretakers gleaned from cues from information that came from signals. - difference in sadness or a difference in facial sadness?

@@

Consistent with signaling models which emphasize the role of conflicts of interest in favoring the use of costlier signals, conflict and signaling behavior were highly correlated. Although conflicts of interest do not always lead to behavioral conflict [@altmann1980; @bateson1994], it is likely many, if not most, conflicts are ultimately caused by conflicts of interest, potentially making it a useful proxy for the underlying conflicts of interest that may increase the benefits of deception and the costs of being deceived\[\[5\]\]. If this is the case, the correlation between signaling and conflict may represent children's attempts at credibly signaling their value of additional investment or the costliness of punishment [@syme2019] to parents who would have good reason to doubt the trustworthiness of simple verbal requests (or the trustworthiness of facial sadness and crying for those resorting to temper tantrums). However, it is also likely that parents view aversive forms of child signaling as conflict in and of itself.

@@

We found evidence of an interaction between how often a child cared for their siblings and the sex of the child in our models examining signal frequencies and cost, whereby there was a positive relationship between the frequency and summed costliness of child signals of need and alloparenting effort for girls but no relationship or a slightly negative one for boys. In many societies, girls are often expected to do more alloparenting within the household and often end up doing so [@helfrecht2016 @weisner1977; @barry1971; @turke1988]. Our models predict that girls in this sample are more likely to provide allocare (confirm), something which matches findins from focus groups of Utilian mothers. If this difference was the outcome of caretakers pressuring girls to do more or resulted in the expectation of more alloparenting, the increase in signaling with alloparenting effort seen among girls may be the result of a strategy to signal the costliness of their efforts in order to invest more in romantic relationships and status seeking behaviors outside the household.

@@

However, child alloparenting effort was negatively associated with caregiver reports of child conflict. One potential reason for this is since parents presumably value the child alloparenting within the family, they see their relationships with children who invest more in the household as less conflictual. In line with this reasoning, we found evidence of a child alloparenting effort by child age interaction, with our model predicting increased levels of conflict for older children with greater alloparenting. Ideally, we could control for signaling behavior to test this. However, the bidirectional relationship between child signaling and conflict with caretakers prevents this (explain better). It is also possible that our hypothesis about conflict over investment in the household is incorrect.

@@

@@

>>>>

As seen in Western cultures [@zeifman2013; @potegal2003; @koch2003] (check for others), the frequencies of crying and temper tantrums decreased with age, as did the frequency of child sadness, summed signal frequency, and signal cost. One possible reason for this is that as children mature, they are more capable of fulfilling their own needs, more able to ask others for what they need, and are less vulnerable to threats. Another is that since one's reproductive value increases as one survives periods of high mortality risk in childhood, the intensity of conflicts of interest with parents decreases since these children are likely to reach adulthood and require less investment to reach adulthood than younger children, current or future [@daly1995]. For this reason, less costly signals may be more likely to work to increase investment, all else being equal\[\[4\]\]. That said, all else may often not be equal if investment in older children is costlier, something which may be reflected in caretakers seeing older children as relatively needy. The decrease in crying and tantrum frequencies with age may also result from the potential for greater reputational costs, something which may also increase the credibility of such signals among older children as previously mentioned.

The effect of age was smallest on child sadness. However, interpretation of this finding is more difficult due to the fact that, unlike crying and temper tantrums, sadness is thought to have important non-signaling functions including drawing one’s attention to a problem, motivating deeper analysis of one’s situation, and causing withdrawal from situations with low fitness prospects [@nesse2019; @nettle2009; @allen2003; @hagen2011]. By asking about sadness as a whole, our measure was unable to disentangle information caretakers gleaned from cues from information that came from signals.

@@

Consistent with signaling models which emphasize the role of conflicts of interest in favoring the use of costlier signals, conflict and signaling behavior were highly correlated. Although conflicts of interest do not always lead to behavioral conflict [@altmann1980; @bateson1994], it is likely many, if not most, conflicts are ultimately caused by conflicts of interest, potentially making it a useful proxy for the underlying conflicts of interest that may increase the benefits of deception for children and the benefits of greater skepticism for parents\[\[5\]\]. If this is the case, the correlation between signaling and conflict may represent children's attempts at credibly signaling their value of additional investment or the costliness of punishment [@syme2019] to parents who would have good reason to doubt the trustworthiness of simple verbal requests (or the trustworthiness of facial sadness and crying for those resorting to temper tantrums). However, it is also likely that parents find some forms of child signaling aversive, and therefore see it as conflict in and of itself.

@@

We found evidence of an interaction between how often a child cared for their siblings and the sex of the child in our lasso models examining signal frequencies and cost, whereby there was a positive relationship between the frequency and summed costliness of child signals of need and alloparenting effort for girls but no relationship or a slightly negative one for boys. In many societies, girls do more alloparenting than boys [@helfrecht2016 @weisner1977; @barry1971; @turke1988], something which matches findings from focus groups of Utilian mothers [add info from an analysis here?]. If this difference was the outcome of caretakers pressuring girls to do more or resulted in the expectation of more alloparenting, the increase in signaling with alloparenting effort seen among girls may be the result of a strategy to signal the costliness of their efforts in order to invest more in romantic relationships and status seeking behaviors outside the household.

@@

[delete based on lasso coefficients?]
However, child alloparenting effort was negatively associated with caregiver reports of child conflict. One potential reason for this is since parents presumably value the child alloparenting within the family, they see their relationships with children who invest more in the household as less conflictual. In line with this reasoning, we found evidence of a child alloparenting effort by child age interaction, with our model predicting increased levels of conflict for older children with greater alloparenting. Ideally, we could control for signaling behavior to test this. However, the bidirectional relationship between child signaling and conflict with caretakers prevents this (explain better). It is also possible that our hypothesis about conflict over investment in the household is incorrect.

[delete now?]
*Body fat discussion*

@@

[delete?]
*In most models, there is a negative association between education and child signaling?*

@@

*Demographic effects*

Outside of the sex by alloparenting effort interaction, we found little evidence of sex differences in child signaling behavior. As adults, females cry more than males, a finding that has been reported across cultures [@vanhemert2011; @becht2001]. One reason why our study may have failed to find significant effects of sex is that sex differences in the causes and reputational effects of signaling need are small or nonexistent until adolescence and young adulthood, with our small sample size for this age range preventing us from detecting a real effect should it exist. It is also possible that the sex difference seen in rates of crying in adults may reflect the lesser ability of females to bargain for better treatment through aggression or the threat of it when faced with adversity [@hagen2016]. Such a sex difference may not exist among preadolescencents due to a lack of difference in strength or athletic ability before puberty [@handelsman2018] or due to their relative lack of formidability compared to adults.

Monthly income had a modest effect on crying frequency and summed signal frequency in our lasso regressions. Many models designed to explain signals of need (e.g., the begging of young birds) predict that offspring who are worse off or are receiving less investment have more to gain from signaling than those with more [@mock2011; @godfray1991; @caro2016a; @godfray2000]. However, it is possible that this can be true and household level resources have little influence on child signaling frequencies between families. As signaling behavior is costly to child, caretakers, and siblings, children may gain direct and indirect fitness benefits from limiting their signaling behavior in times of resource stress (i.e., when the chances of gaining investment or the size of investment are small). Similarly, children in high resource families may also be motivated to signal at high frequencies due to the potential for larger pieces of investment. We know of no empirical explorations of this in humans (check more and confirm). However, multiple game theoretical models examining parent-offspring conflict have found that parental resources have a modest impact on the evolutionary stable value for offspring demand for resources [@bossan2013; @mock1997 [confirm]]. Compared to overall effort in eliciting need, resource availability may have a larger impact on how children signal for more investment based on the finding that birds in stable and unusually favorable environments are more likely to compete with siblings through signals thought to reveal information about need, while those in relatively unpredictable and unusually poor environments have been found to compete more through signals thought to reveal information about quality [@caro2016a].

Despite the differences in wealth, job certainty, discrimination, and living conditions [@davis2014; @garcia2017; @garcia2018; @stonich2000] between recent immigrants from the mainland and native Utilians, primary language, immigration status, and residence in Camponado did not predict child signaling in our lasso regressions. However, neighborhood quality was preserved in the lasso regressions predicting every signaling measure but sadness and was among the variables with the largest negative coefficients. One reason for this might that neighborhood quality affects the amount of adversity children face, thereby increasing their neediness [no evidence of this when neighborhood quality is included in models predicting relative child need. ]

[old]
Despite population level differences in wealth, job certainty, discrimination, and living conditions [@davis2014; @garcia2017; @garcia2018; @stonich2000], we generally did not find an effect of population on child signaling behavior. Although we sought data from both English-speaking participants born on the island and Spanish-speaking immigrants from the mainland, our smaller sample of English-speaking Utilians may have limited our ability to detect real differences between the populations. It is possible that differences in wealth and status between the groups translate to differences in child signaling due to children facing different types adversity or weighting different types of investment differently. It is also likely that the two groups differ in how caretakers could respond to child signaling, and caretakers between the two populations may also differ in their cultural models of parenting. However, as previously mentioned, it is conceivable that household level differences have little effect on child signaling behavior.

possessions
